---
title: "ESM 206 Assignment 4 Task 2 Bonanza Creek"
author: "Sarah Hamilton"
date: "11/6/2021"
output: 
  html_document: 
    toc: yes
    theme: cerulean
    number_sections: yes
    code_folding: hide
---

# Overview

This section contains an overview of what is contained in this report. 

## Introduction

This report provides an exploratory analysis of juvenile snowshoe hares in the Bonanza Creek Long Term Ecological Research Site (Bonanza Creek LTER, 2021). The report first explores total counts of juvenile hare trappings over the years 1999 to 2012. It then explores juvenile hare weight by sex at three different sites at Bonanza Creek where juvenile hares were found. Finally, it explores the relationship between juvenile hare hind foot length and weight. 

## Data and Analysis

The data used in this report comes from the Bonanza Creek Long Term Ecological Research Site, which is run by the University of Alaska Fairbanks (Bonanza Creek LTER, 2021). For this analysis, only juvenile hares were included. In section 2.1, the annual juvenile hare trap counts were plotted over time on a histogram. In section 2.2, a beeswarm plot was used to visualize the weight of juvenile hares at each site of Bonanza Creek. In section 2.3, a basic summary statistics table was made to clearly compare the weights of female and male juvenile hares, and Cohen's d effect size test and a two-sample unpaired t-test were run to further compare the weights. In section 2.4, a linear regression model was run on a scatterplot of weight versus hind foot length to determine the relationship between the two variables, and a Pearson's r correlation test was also run to further investigate the correlation between these two variables. 


```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
#Install the necessary packages 
library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(kableExtra)
library(ggbeeswarm)
library(effsize)
library(broom)
library(equatiomatic)
```

# Juvenile Hare Investigation

Analyses of juvenile hare trap counts at different sites, juvenile hare weights between sexes, and the relationship between juvenile weights and hind foot length are included below. 

## Annual Juvenile Hare Trap Counts

```{r fig.align='center'}
#Read in the data
hares_raw <- read_csv(here("data","bonanza_hares.csv"))

#Filter for juveniles and add a new date column and a year column using functions from lubridate
hares <- hares_raw %>%
  filter(age == "j") %>% #filter to only include juveniles
  mutate(date_new = mdy(date)) %>% #make a new date column
  mutate(year_new = year(date_new)) #add a new year column

#Group the data by year using group_by(), then summarize it to get the counts using summarize()
hares_annual_counts <- hares %>%
  group_by(year_new) %>% #group by year
  summarize(count_hares = n()) #get the sample size

#Calculate summary statistics to reference in text below
min_count <- min(hares_annual_counts$count_hares) #calculate the minimum count
max_count <- max(hares_annual_counts$count_hares) #calculate the maximum count
mean_count <- round(mean(hares_annual_counts$count_hares)) #calculate the mean count
median_count <- median(hares_annual_counts$count_hares) #calculate the median count

#Plot the count of hares by year using ggplot() and geom_col()
ggplot(data = hares_annual_counts, aes(x = year_new, y = count_hares)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Year", 
       y = "Number of Juvenile Hare Trappings", 
       caption = "Figure 1. Juvenile Hare Trappings Over Time. The x axis shows the year of data collection, \n and the y axis shows the total number of hare trappings. Data: Bonanza Creek LTER, 2021") +
  theme(plot.caption = element_text(hjust=0.5)) #center the caption

```

Figure 1 shows peaks in annual juvenile hare trappings around 1999, 2005, and 2008 with a large dip in trappings occurring around 2002 and 2009. The minimum number of trappings in a given year was `r min_count` and the maximum number of trappings was `r max_count`. The average number of trappings per year is `r mean_count`, and the median number of trappings per year is `r median_count`. There seem to be more consistently higher numbers of trappings in the late 1990s than there are in later years. In order to standardize juvenile hare population for future analyses and ensure that these observations are consistent with the actual hare population, I would suggest that future data collection occurs on the same days each year with the same number of traps on each day to ensure that the number of trappings per year are calculated in a more uniform way. 

## Juvenile Hare Weights

```{r fig.align='center'}
#Create a plot of juvenile hare weights by sex and grid 
hares_updated_grid <- hares %>% #use case_when() to rename the sites to be more clear
  mutate(grid_new = case_when(
    grid == "bonbs" ~ "Bonanza Black Spruce",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian"
  )) %>%
  mutate(sex_new = case_when( #use case_when() to update the sex names for the table
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",
    sex == "NA" ~ "Not Recorded"
  ))


ggplot(data = hares_updated_grid,aes(x = sex_new, y = weight)) + #create a beeswarm plot
  geom_beeswarm(aes(color = grid_new), cex=2, show.legend = FALSE) + #color by sex and spread out using cex
  theme_bw() + #add a theme
  labs(x = "Sex", 
       y = "Weight (g)",  
       caption = "Figure 2. Juvenile Hare Weight by Site and Sex. The sex of the observations collected are plotted \n on the x axis, with the weight in grams plotted on the y axis. \n The colors indicate the site of the trapped hare. Data: Bonanza Creek LTER, 2021") + #add labels and caption
  theme(plot.caption = element_text(hjust=0.5)) + #center the caption
  facet_wrap(~grid_new)
```


```{r fig.align='center'}
#Create a table of summary statistics of juvenile hare weights by sex and grid
hare_weight_summary <- hares_updated_grid %>%
  group_by(grid_new, sex_new) %>% #group the data by grid/site and sex
  summarize(
    mean_weight = round(mean(weight, na.rm = TRUE),2), #calculate mean of weight
    sd_weight = round(sd(weight, na.rm = TRUE),2), #calculate standard deviation of weight
    count_weight = n() #calculate sample size
  )

hare_weight_summary %>% #make the table easier to read using kable()
  kable(col.names = c("Site","Sex","Average Weight (g)","Standard Deviation of Weight (g)","Sample Size"),#rename the columns
        caption = "Table 1. Summary statistics for the weights of male and female hares by site. Data: Bonanza Creek LTER, 2021") %>% #add a table caption
  kable_styling(bootstrap_options = "striped", full_width = FALSE) #set visual preferences 
```

The Bonanza Black Spruce site seems to have higher average weights than the Bonanza Mature and Bonanza Riparian sites, though this cannot be confirmed without further testing of the significance of these differences. This is evident in the summary statistics table, as well as by the shape of the beeswarm distributions, which are more heavily weighted to higher weights for the Bonanza Black Spruce site. The other two sites have very similar average weights. However, the Bonanza Black Spruce site has the lowest sample size, so it is possible that the values for this site differ partially because of the smaller sample. The data also seems to indicate higher weights in male juvenile hares than in female juvenile hares, as the average male weights are consistently higher than the female weights at each site. 

## Juvenile Weight Comparison

```{r}
#Filter out NA values because we are comparing males to females 
hares_mf <- hares %>% 
  filter(sex != "NA") %>% #filter to remove NAs
  mutate(sex_new = case_when( #use case_when() to update the sex names for the table
    sex == "f" ~ "Female",
    sex == "m" ~ "Male"))

#Create a table comparing means of male and female juvenile hares
hares_summary <- hares_mf %>%
  group_by(sex_new) %>% #group by sex
  summarize(mean_weight2 = round(mean(weight, na.rm = TRUE),2), #calculate mean weight
            sd_weight2 = round(sd(weight, na.rm = TRUE),2), #calculate std. dev. weight
            count2 = n()) #calculate sample sizes
```

```{r include = FALSE}
#Exploratory data analysis to see if male and female weights are normally distributed
weight_hist <- ggplot(data = hares, aes(x = weight))+
  geom_bar(aes(color = sex))+
  facet_wrap(~sex)
weight_hist #outputs the weight_hist plot
#The histogram looks relatively normal, indicating that a t-test is probably appropriate.

weight_qq <- ggplot(data = hares_mf) +
  geom_qq(aes(sample = weight, color = sex))
weight_qq #outputs the weight_qq plot
#The qq plot results in mostly linear lines for both sexes, indicating that a t-test is probably appropriate.
```

```{r fig.align='center'}
#Producing organized table of information
hares_summary %>% kable(
  col.names = c("Sex","Average Weight (g)","Standard Deviation of Weight (g)","Sample Size"), #change column names
  caption = "Table 2. Summary statistics for male and female juvenile hares. Data: Bonanza Creek LTER, 2021") %>% #add caption
  kable_styling(bootstrap_options = "striped", full_width = FALSE) #set visual preferences 
```

```{r}
#Calculating statistics to reference in report
#1. difference in means 
dif_means <- hares_summary$mean_weight2[2]-hares_summary$mean_weight2[1]

#2. effect size
female_weights <- hares_mf %>%
  filter(sex == "f") %>% #filter for female
  filter(weight != "NA") %>% #filter out NA values
  pull(weight) #pull out weight column

male_weights <- hares_mf %>%
  filter(sex == "m") %>% #filter for male
  filter(weight != "NA") %>% #filter out NA values
  pull(weight) #pull out weight column

weight_cohen <- cohen.d(female_weights, male_weights) #run cohen effect size test
#weight_cohen #output cohen effect size 

#3. outcome of unpaired two sample t-test
weight_t <- t.test(female_weights, male_weights) #run unpaired two sample t test
#weight_t #output t test results
  
```

The mean weight for male juvenile hares (`r hares_summary$mean_weight2[2]` g $\pm$ `r hares_summary$sd_weight2[2]` g) is `r dif_means` g greater than that for female juvenile hares (`r hares_summary$mean_weight2[1]` g $\pm$ `r hares_summary$sd_weight2[1]` g). The effect size for the means is small (Cohen's d = `r round(weight_cohen$estimate,2)`) and the difference in means is significant (two-sample t-test, t = `r round(weight_t$statistic,2)`, p = `r round(weight_t$p.value,3)`).

## Juvenile Weight and Hind Foot Length

```{r fig.align='center'}
ggplot(data = hares, aes(x = hindft, y = weight)) + #make a plot of weight vs. hind foot length
  geom_point() + #scatter plot
  theme_bw() + #add a theme
  geom_smooth(method = "lm") + #add a linear regression line
  labs(x = "Hind Foot Length (mm)", y = "Weight (g)",
       caption = "Figure 3. Juvenile hare weight versus hind foot length. The juvenile hares' hind foot length is plotted on the x axis \n in millimeters, and their weights in grams are plotted on the y axis. Data: Bonanza Creek LTER, 2021") + #add labels and a caption
  theme(plot.caption = element_text(hjust=0.5)) #center the caption
```

```{r}
#Find linear regression model
hare_lm <-lm(weight ~ hindft, data = hares) #find linear regression model for weight as a function of hind foot length
#hare_lm #output lm model

#Get tidy output of model
hare_lm_tidy <- tidy(hare_lm) #get a tidy output of lm()
#hare_lm_tidy #output lm tidy model

#Get output of model 
hare_lm_glance <- glance(hare_lm) #get output of model fit and statistical significance
#hare_lm_glance #output lm glance 

#Output thelinear regression model 
extract_eq(model = hare_lm, use_coefs = TRUE) #return the model equation in LaTeX

#Get Pearson's R correlation
hare_pearson_test <- cor.test(hares$hindft, hares$weight) 
hares_pearson_r <- hare_pearson_test$estimate #pull out the pearson's r value
hare_pearson_tidy <- tidy(hare_pearson_test)
```

The output of this linear regression indicates that on average, for every 1 mm increase in hind foot length, weight increases by `r round(hare_lm$coefficients[2],1)` g. The y intercept of the formula does not have any inherent meaning in this case. This regression model has an R^2^ value of `r round(hare_lm_glance$r.squared,3)` (p < 0.001), which means that `r 100*round(hare_lm_glance$r.squared,3)`% of the variance in weight is predictable from hind foot length. The Pearson's r correlation for the model is `r round(hares_pearson_r,2)` (p < 0.001). indicating that the weight of a hare is moderately influenced by its hind foot length. The assumption of homoscedasticity may be a concern for this linear regression model because the residuals appear to be smaller at first and then larger as hind foot length increases, indicating inconsistent residuals. Therefore, it is possible that these variables are slightly linearly related, but there is also reason to be skeptical of their dependence due to the low R^2^ and moderate Pearson's r correlation values, and the limitations to the linear regression model.

# Summary

This analysis resulted in the following findings:

- The total number of juvenile hare trappings fluctuated greatly over time, peaking in 1999, 2005, and 2011. The average number of juvenile hare trappings per year is 32.

- The Bonanza Black Spruce site had the highest average juvenile hare weight.

- Male juvenile hares consistently had higher average weights than female juvenile hares at every site. The mean weight for males was 90 grams greater than that for female hares.

- A regression model for weight versus hind foot length indicated that about 30% of the variance in weight is predictable from hind foot length, and a Pearson's r correlation test indicated weak correlation between the two variables. 

# Citations

Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14





