---
title: "ESM 206 Assignment 4 Task 2 Bonanza Creek"
author: "Sarah Hamilton"
date: "11/6/2021"
output: 
  html_document: 
    toc: yes
    theme: cerulean
    number_sections: yes
    code_folding: hide
---

# Introduction

DO THIS STILL******* A 4 - 5 sentence "mini-introduction" to what this study is exploring, including basic background information about the subject, site and study (with citations as needed), and a brief overview of what is included in the exploratory report. You should probably do this last so you know what’s contained in the report.

# Data and Analysis

DO THIS STILL******* A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.


```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(kableExtra)
library(ggbeeswarm)
library(effsize)
library(broom)
library(equatiomatic)
```

# Juvenile Hare Investigation

In this section, I conduct analyses of juvenile hare trap counts at different sites, juvenile hare weights between sexes, and the relationship between juvenile weights and hind foot length. 

## Annual Juvenile Hare Trap Counts

```{r}
#Read in the data
hares_raw <- read_csv(here("data","bonanza_hares.csv"))

#Filter for juveniles and add a new date column and a year column using functions from lubridate
hares <- hares_raw %>%
  filter(age == "j") %>% #filter to only include juveniles
  mutate(date_new = mdy(date)) %>% #make a new date column
  mutate(year_new = year(date_new)) #add a new year column

#Group the data by year using group_by(), then summarize it to get the counts using summarize()
hares_annual_counts <- hares %>%
  group_by(year_new) %>% #group by year
  summarize(count_hares = n()) #get the sample size

#Calculate summary statistics to reference in text below
min_count <- min(hares_annual_counts$count_hares) #calculate the minimum count
max_count <- max(hares_annual_counts$count_hares) #calculate the maximum count
mean_count <- round(mean(hares_annual_counts$count_hares)) #calculate the mean count
median_count <- median(hares_annual_counts$count_hares) #calculate the median count

#Plot the count of hares by year using ggplot() and geom_col()
ggplot(data = hares_annual_counts, aes(x = year_new, y = count_hares)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Year", 
       y = "Number of Juvenile Hare Trappings", 
       title = "Juvenile Hare Trappings Over Time",
       caption = "Figure 1. Juvenile Hare Trappings Over Time. The x axis shows the year of data collection, and the y axis shows \n the total number of hare trappings.")

```

Figure 1 shows peaks in annual juvenile hare trappings around 1999, 2005, and 2008 with a large dip in trappings occurring around 2002 and 2009. The minimum number of trappings in a given year was `r min_count` and the maximum number of trappings was `r max_count`. The average number of trappings per year is `r mean_count`, and the median number of trappings per year is `r median_count`. There seem to be more consistently higher numbers of trappings in the late 1990s than there are in later years. In order to standardize juvenile hare population for future analyses and ensure that these observations are consistent with the actual hare population, I would suggest that future data collection occurs on the same days each year with the same number of traps on each day to ensure that the number of trappings per year are calculated in a more uniform way. 

## Juvenile Hare Weights

```{r}
#Create a plot of juvenile hare weights by sex and grid 
hares_updated_grid <- hares %>% #use case_when() to rename the sites to be more clear
  mutate(grid_new = case_when(
    grid == "bonbs" ~ "Bonanza Black Spruce",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian"
  ))

ggplot(data = hares_updated_grid,aes(x = grid_new, y = weight)) + #create a beeswarm plot
  geom_beeswarm(aes(color = sex), cex=2) + #color by sex and spread out using cex
  theme_bw() + #add a theme
  scale_color_discrete(labels = c("Female","Male","Not Recorded")) + #update the legend labels to be more clear
  labs(x = "Site", 
       y = "Weight (g)",  
       title = "Juvenile Hare Weight by Site", 
       color = "Sex",
       caption = "Figure 2. Juvenile Hare Weight by Site. The site at which the observations were collected is plotted on the x \n axis, with the weight in grams plotted on the y axis. The colors indicate the sex of the trapped hare, with red indicating \n a female hare, blue indicating a male hare, and gray indicating that the sex of the hare was not recorded.") #add labels and caption
```


```{r}
#Create a table of summary statistics of juvenile hare weights by sex and grid
hare_weight_summary <- hares_updated_grid %>%
  mutate(sex_new = case_when( #use case_when() to update the sex names for the table
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",
    sex == "NA" ~ "Not Recorded"
  )) %>%
  group_by(grid_new, sex_new) %>% #group the data by grid/site and sex
  summarize(
    mean_weight = round(mean(weight, na.rm = TRUE),2), #calculate mean of weight
    sd_weight = round(sd(weight, na.rm = TRUE),2), #calculate standard deviation of weight
    count_weight = n() #calculate sample size
  )

hare_weight_summary %>% #make the table easier to read using kable()
  kable(col.names = c("Site","Sex","Average Weight (g)","Standard Deviation of Weight (g)","Sample Size"),#rename the columns
        caption = "Table 1. Summary statistics for the weights of male and female hares by site.") %>% #add a table caption
  kable_styling(bootstrap_options = "striped", full_width = FALSE) #set visual preferences 
```

The Bonanza Black Spruce site has higher average weights than the Bonanza Mature and Bonanza Riparian sites. This is evident in the summary statistics table, as well as by the shape of the beeswarm distributions, which are more heavily weighted to higher weights for the Bonanza Black Spruce site. The other two sites have very similar average weights. However, the Bonanza Black Spruce site has the lowest sample size, so it is possible that the values for this site differ partially because of the smaller sample. The data also seems to indicate higher weights in male juvenile hares than in female juvenile hares, as the average male weights are consistently higher than the female weights at each site. 

## Juvenile Weight Comparison

```{r}
#Filter out NA values because we are comparing males to females 
hares_mf <- hares %>% 
  filter(sex != "NA") %>% #filter to remove NAs
  mutate(sex_new = case_when( #use case_when() to update the sex names for the table
    sex == "f" ~ "Female",
    sex == "m" ~ "Male"))

#Create a table comparing means of male and female juvenile hares
hares_summary <- hares_mf %>%
  group_by(sex_new) %>% #group by sex
  summarize(mean_weight2 = round(mean(weight, na.rm = TRUE),2), #calculate mean weight
            sd_weight2 = round(sd(weight, na.rm = TRUE),2), #calculate std. dev. weight
            count2 = n()) #calculate sample sizes
```

```{r include = FALSE}
#Exploratory data analysis to see if male and female weights are normally distributed
weight_hist <- ggplot(data = hares, aes(x = weight))+
  geom_bar(aes(color = sex))+
  facet_wrap(~sex)
weight_hist #outputs the weight_hist plot
#The histogram looks relatively normal, indicating that a t-test is probably appropriate.

weight_qq <- ggplot(data = hares_mf) +
  geom_qq(aes(sample = weight, color = sex))
weight_qq #outputs the weight_qq plot
#The qq plot results in mostly linear lines for both sexes, indicating that a t-test is probably appropriate.
```

```{r}
#Producing organized table of information
hares_summary %>% kable(
  col.names = c("Sex","Average Weight (g)","Standard Deviation of Weight (g)","Sample Size"), #change column names
  caption = "Table 2. Summary statistics for male and female juvenile hares.") %>% #add caption
  kable_styling(bootstrap_options = "striped", full_width = FALSE) #set visual preferences 
```

```{r}
#Calculating statistics to reference in report
#1. difference in means 
dif_means <- hares_summary$mean_weight2[2]-hares_summary$mean_weight2[1]

#2. effect size
female_weights <- hares_mf %>%
  filter(sex == "f") %>% #filter for female
  filter(weight != "NA") %>% #filter out NA values
  pull(weight) #pull out weight column

male_weights <- hares_mf %>%
  filter(sex == "m") %>% #filter for male
  filter(weight != "NA") %>% #filter out NA values
  pull(weight) #pull out weight column

weight_cohen <- cohen.d(female_weights, male_weights) #run cohen effect size test
weight_cohen #output cohen effect size 

#3. outcome of unpaired two sample t-test
weight_t <- t.test(female_weights, male_weights) #run unpaired two sample t test
weight_t #output t test results
  
```

The mean weight for male juvenile hares (`r hares_summary$mean_weight2[2]` g) is `r dif_means` g greater than that for female juvenile hares (`r hares_summary$mean_weight2[2]` g). The effect size for the means is small (Cohen's d = `r round(weight_cohen$estimate,2)`) and the difference in means is significant (two-sample t-test, t = `r round(weight_t$statistic,2)`, p = `r round(weight_t$p.value,3)`).

## Juvenile Weight and Hind Foot Length

```{r}
ggplot(data = hares, aes(x = hindft, y = weight)) + #make a plot of weight vs. hind foot length
  geom_point() + #scatter plot
  theme_bw() + #add a theme
  geom_smooth(method = "lm") + #add a linear regression line
  labs(x = "Hind Foot Length (mm)", y = "Weight (g)", title = "Juvenile Hare Weight versus Hind Foot Length",
       caption = "Figure 3. Juvenile hare weight versus hind foot length. The juvenile hares' hind foot length \n is plotted on the x axis in millimeters, and their weights in grams are plotted on the y axis.") #add labels and a caption
```

```{r}
hare_lm <-lm(weight ~ hindft, data = hares) #find linear regression model for weight as a function of hind foot length
#hare_lm #output lm model

hare_lm_tidy <- tidy(hare_lm) #get a tidy output of lm()
#hare_lm_tidy #output lm tidy model

hare_lm_glance <- glance(hare_lm) #get output of model fit and statistical significance
#hare_lm_glance #output lm glance 

extract_eq(model = hare_lm, use_coefs = TRUE) #return the model equation in LaTeX
```

The output of this linear regression indicates that on average, for every 1 mm increase in hind foot length, weight increases by `r round(hare_lm$coefficients[2],1)` g. This regression model has an R^2 value of `r round(hare_lm_glance$r.squared,3)`, which means that `r 100*round(hare_lm_glance$r.squared,3)`% of the variance in weight is predictable from hind foot length. The Pearson's r correlation for the model is ________. The assumption of _____ may be a concern for this linear regression model because _________.

# Summary

DO THIS STILL***** Include a very short summary describing your main findings from your juvenile hares exploration (bullet points are fine)

# Citations

Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14





