---
title: "ESM 206 Assignment 4 Task 2 Bonanza Creek"
author: "Sarah Hamilton"
date: "11/6/2021"
output: 
  html_document: 
    toc: yes
    theme: cerulean
    number_sections: yes
    code_folding: hide
---

# Introduction

DO THIS STILL******* A 4 - 5 sentence "mini-introduction" to what this study is exploring, including basic background information about the subject, site and study (with citations as needed), and a brief overview of what is included in the exploratory report. You should probably do this last so you know what’s contained in the report.

# Data and Analysis

DO THIS STILL******* A 3 - 4 sentence section in which you briefly describe the data, including citation & credit as needed, and provide an overview of the main analysis, statistical methods and tools used.

Data Source: Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative. https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14


```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(janitor)
library(here)
library(lubridate)
library(kableExtra)
library(ggbeeswarm)
```

# Juvenile Hare Investigation

In this section, I ...

## Annual Juvenile Hare Trap Counts

```{r}
#Read in the data
hares_raw <- read_csv(here("data","bonanza_hares.csv"))

#Add a new date column and a year column using functions from lubridate
hares <- hares_raw %>%
  mutate(date_new = mdy(date)) %>%
  mutate(year_new = year(date_new))

#Group the data by year using group_by(), then summarize it to get the counts using summarize()
hares_annual_counts <- hares %>%
  group_by(year_new) %>%
  summarize(count_hares = n())

#Calculate summary statistics to reference in text below
min_count <- min(hares_annual_counts$count_hares) #calculate the minimum count
max_count <- max(hares_annual_counts$count_hares) #calculate the maximum count
mean_count <- round(mean(hares_annual_counts$count_hares)) #calculate the mean count
median_count <- median(hares_annual_counts$count_hares) #calculate the median count

#Plot the count of hares by year using ggplot() and geom_col()
ggplot(data = hares_annual_counts, aes(x = year_new, y = count_hares)) +
  geom_col() +
  theme_minimal() +
  labs(x = "Year", 
       y = "Number of Juvenile Hare Trappings", 
       title = "Juvenile Hare Trappings Over Time",
       caption = "Figure 1. Juvenile Hare Trappings Over Time. The x axis shows the year of data collection, and the y axis shows \n the total number of hare trappings.")

```

Figure 1 shows peaks in annual hare trappings around 1999 and 2009, with a large dip in trappings occurring around 2002. The minimum number of trappings in a given year was `r min_count` and the maximum number of trappings was `r max_count`. The average number of trappings per year is `r mean_count`, and the median number of trappings per year is `r median_count`. There seem to be more consistently higher numbers of trappings in the late 2000s and early 2010s than there are in earlier years. In order to standardize juvenile hare population for future analyses and ensure that these observationsa are consistent with the actual hare population, I would suggest that future data collection occurs on the same days each year with the same number of traps on each day to ensure that the number of trappings per year are calculated in a more uniform way. 

## Juvenile Hare Weights

```{r}
#Create a plot of juvenile hare weights by sex and grid 
hares_updated_grid <- hares %>% #use case_when() to rename the sites to be more clear
  mutate(grid_new = case_when(
    grid == "bonbs" ~ "Bonanza Black Spruce",
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian"
  ))

ggplot(data = hares_updated_grid,aes(x = grid_new, y = weight)) + #create a beeswarm plot
  geom_beeswarm(aes(color = sex), cex=0.75) + #color by sex and spread out using cex
  theme_bw() + #add a theme
  scale_color_discrete(labels = c("Female","Male","Not Recorded")) + #update the legend labels to be more clear
  labs(x = "Site", 
       y = "Weight (g)",  
       title = "Juvenile Hare Weight by Site", 
       color = "Sex",
       caption = "Figure 2. Juvenile Hare Weight by Site. The site at which the observations were collected is plotted on the x \n axis, with the weight in grams plotted on the y axis. The colors indicate the sex of the trapped hare, with red indicating \n a female hare, blue indicating a male hare, and gray indicating that the sex of the hare was not recorded.") #add labels and caption
```


```{r}
#Create a table of summary statistics of juvenile hare weights by sex and grid
hare_weight_summary <- hares_updated_grid %>%
  mutate(sex_new = case_when( #use case_when() to update the sex names for the table
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",
    sex == "NA" ~ "Not Recorded"
  )) %>%
  group_by(grid_new, sex_new) %>% #group the data by grid/site and sex
  summarize(
    mean_weight = round(mean(weight, na.rm = TRUE),2), #calculate mean of weight
    sd_weight = round(sd(weight, na.rm = TRUE),2) #calculate standard deviation of weight
  )

hare_weight_summary %>% #make the table easier to read using kable()
  kable(col.names = c("Site","Sex","Average Weight (g)","Standard Deviation of Weight (g)"),#rename the columns
        caption = "Table 1. Summary statistics for the weights of male and female hares by site.") %>% #add a table caption
  kable_styling(bootstrap_options = "striped", full_width = FALSE) #set visual preferences 
```

The Bonanza Black Spruce and Bonanza Riparian sites have higher average weights than Bonanza Mature. This is evident in the summary statistics table, as well as by the shape of the beeswarm distributions, which are more heavily weighted to higher weights for the Bonanza Black Sprice and Bonanza Riparian sites. There do not seem to be large differences in weights between male and female hares, as the difference between the average male and female weight at every site is fewer than 80 grams. 

## Juvenile Weight Comparison



## Juvenile Weight and Hind Foot Length

# Summary

# Citations





